---
title: "K-means Metaclustering for MxIF"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged 
    geometry: margin=2cm
    highlight: textmate
    theme: journal
    fig_crop: false
    toc: true
    toc_float: true
  pdf_document: default
---
<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

requiredPackages <- c("data.table", "plyr", "dplyr", "knitr", "ComplexHeatmap", "ggplot2", "pander")
for (package in requiredPackages) { #Installs packages if not yet installed
  if (!requireNamespace(package, quietly = TRUE))
  install.packages(package)
}

library(data.table)
library(plyr)
library(dplyr)
library(knitr)
library(ComplexHeatmap)
library(ggplot2)
library(pander)

set.seed(123)
```

Be sure to check the distribution of ROIs across metaclusters to verify that the ROIs are approximately evenly represented within the metaclusters. Unbalanced ROI proportions indicate unresolved batch effects between ROIs. **If there are more than 15 ROIs, plot legends will be omitted as they become uninterpretable.**

```{r}
allmarker_files <- list.files('.', pattern = "^all_markers_clean_*", full.names = T)
cluster_files <- list.files('.', pattern = "^scimap_clusters_.*.csv", full.names = T)
configs <- fread("configs.csv", sep=",")
marker_configs <- fread("marker_configs.csv", sep=",")

roi_df <- fread(allmarker_files[1])


# n_metaclusters <- as.numeric(configs$value[configs$object == "n_metaclusters"])
min_metaclusters <- as.numeric(configs$value[configs$object == "min_metaclusters"])
max_metaclusters <- as.numeric(configs$value[configs$object == "max_metaclusters"])
metacluster_range <- min_metaclusters:max_metaclusters

# Default marker set
default_markers <- c("CD20", "FOXP3", "CD8", "CD4", "Ecad", "CD3e", "CD68", "CD45", "CD14", "CD31")

## Choose markers for clustering
if(nrow(marker_configs) > 0 & all(sapply(marker_configs$marker, function(x) any(grepl(x, colnames(roi_df)))))) {
  markers_selected <- marker_configs$marker[!grepl("DAPI", marker_configs$marker)]
} else if (all(sapply(default_markers, function(x) any(grepl(x, colnames(roi_df)))))) {
  markers_selected <- default_markers
} else {
  markers_selected <- sapply(grep("Median", colnames(roi_df), value = T), function(x) sub(": Cell: Median", "", x))
  markers_selected <- markers_selected[!grepl("DAPI", markers_selected)]
}
markers_selected <- markers_selected[markers_selected != ""]
```


```{r}
arcsine_transform <- function(x) {
  y <- x / max(x)
  asin(sqrt(y))
}

zscorenorm <- function(marker) {
  mu = mean(marker)
  sd = sd(marker)
  (marker - mu) / sd
}

get_normalized_counts <- function(roi_allmarkers) {
  counts <- fread(roi_allmarkers)
  counts <- counts %>% rename("x" = matches("Centroid X"), "y" = matches("Centroid Y"), "imageid" = "roi")
  counts <- counts %>% select(imageid, all_of(setdiff(colnames(.), "imageid"))) %>% select(-contains("DAPI"))

  cols <- grep("Median", colnames(counts), value = T)

  for (j in cols) set(counts, j = j, value = arcsine_transform(counts[[j]]))
  for (j in cols) set(counts, j = j, value = zscorenorm(counts[[j]]))

  counts
}

## join clusters and quant data
get_normalized_cluster_summary <- function(roi_allmarkers, roi_clusters) {
  normcounts <- get_normalized_counts(roi_allmarkers)
  clusters <- fread(roi_clusters) %>%
    rename("x" = "X_centroid", "y" = "Y_centroid")
  imageid <- clusters$imageid[1]
  
  cluster_medians <- 
    inner_join(clusters, normcounts, by = c("x", "y", "imageid")) %>% 
    filter(kmeans != "Artifact") %>%
    select(x, y, imageid, kmeans, all_of(paste0(markers_selected, ": Cell: Median"))) %>% 
    group_by(kmeans) %>% 
    summarise_at(paste0(markers_selected, ": Cell: Median"), list(mean = mean, median = median, sd = sd), na.rm = T) %>%
    mutate(kmeans = paste0(imageid, "_", kmeans), imageid = imageid)
  
  data.frame(cluster_medians, check.names = F)
}

make_heatmap <- function(dat, n_metaclusters) {
  smSub <- dat %>%
    mutate_all(funs(replace(., is.na(.), 0))) %>%
    ungroup()
  
  # Create matrix to be used in the heatmap
  mat2 <- smSub %>%
    select(-c(kmeans, imageid)) %>% replace(is.na(.), 0) %>%
    # as.matrix() %>% t() %>% pheatmap:::scale_rows()
    as.matrix() %>% pheatmap:::scale_rows() %>% t()
  
  ## Annotation for cluster
  textannot <- if(length(allmarker_files) > 15) {
    rep(" ", length(smSub$kmeans))
    } else {
      smSub$kmeans
    }
  
  ha = HeatmapAnnotation(FOV = smSub$imageid,
                         ClusterID = anno_text(textannot, gp = gpar(fontsize = 8)), 
                         show_legend = ifelse(length(allmarker_files) > 15, FALSE, TRUE))
  
  mat2[is.nan(mat2)] <- 0
  colnames(mat2) <- smSub$cluster
  
  Heatmap(mat2, #col = col_fun,
          name = "mat",
          row_names_gp = gpar(fontsize = 10),
          top_annotation = ha,
          # bottom_annotation = ba,
          column_km = n_metaclusters,
          border = TRUE)
}
```


```{r fig.height=8, fig.width=10}
all_normalized_cluster_summaries <- mapply(get_normalized_cluster_summary, allmarker_files, cluster_files, SIMPLIFY = F)
all_normalized_cluster_summaries <- bind_rows(all_normalized_cluster_summaries, .id = "column_label")
all_normalized_cluster_summaries <- all_normalized_cluster_summaries %>% select(-column_label)

centroids_out <- all_normalized_cluster_summaries %>% select(-imageid)
fwrite(centroids_out, "arcsin_zscore_kmeans_centroids.csv")

all_normalized_cluster_summaries <- all_normalized_cluster_summaries %>% select(!contains("sd") & !contains("_mean"))
colnames(all_normalized_cluster_summaries) <- sapply(colnames(all_normalized_cluster_summaries), function(x) sub("_median", "", x))
```

```{r}
df_list <- lapply(cluster_files, function(x) fread(x))

names(df_list) <- basename(cluster_files)
```


```{r fig.height=8, fig.width=12, results='asis'}
for (i in seq_along(metacluster_range)) {
  
  n_metaclusters <- metacluster_range[i]
  
  pandoc.header(paste(n_metaclusters, "metaclusters"),level=2,style=c("atx","setext"))
  
  set.seed(123)
  
  cat("\n")
  pandoc.header("Heatmap", level = 3)
  
  ht <- draw(make_heatmap(all_normalized_cluster_summaries, n_metaclusters))
  
  dat <- all_normalized_cluster_summaries
  
  smSub <- dat %>%
    mutate_all(funs(replace(., is.na(.), 0))) %>%
    ungroup()
  
  get_metacluster_table <- function(metacluster) {
    metaclust_ind <- unlist(ht@ht_list$mat@column_order_list[[metacluster]])
    dat <- data.frame(metacluster = metacluster, cluster = smSub$kmeans[metaclust_ind], imageid = smSub$imageid[metaclust_ind])
    dat
  }
  
  all_metacluster_counts <- ldply(as.character(1:n_metaclusters), get_metacluster_table)
  
  
  meta_orig_tab <- data.frame(table(all_metacluster_counts$metacluster, all_metacluster_counts$imageid))
  colnames(meta_orig_tab) <- c("metacluster", "ROI", "Freq")
  
  cat("\n")
  pandoc.header("Distribution of ROIs across metaclusters", level = 3)
  
  p <- ggplot(meta_orig_tab, aes(x = factor(metacluster, levels = as.character(1:n_metaclusters)), y = Freq, fill = ROI)) +
    geom_bar(position="fill", stat="identity", show.legend = ifelse(length(allmarker_files) > 15, FALSE, TRUE)) +
    theme_bw() +
    xlab("metacluster")
  
  plot(p)
  cat('\n')
  
  ## Use all_metacluster_counts to map back to coordinates (in cluster files) and write out mapped metacluster file
  all_metacluster_counts$metacluster <- paste0("kmeans_meta_", all_metacluster_counts$metacluster)
  
  for (j in cluster_files) {
    single_df <- df_list[[basename(j)]]
    single_df$cluster <- paste(single_df$imageid, single_df$kmeans, sep = "_")
    meta_mapped <- left_join(single_df, all_metacluster_counts, by = c("cluster" = "cluster", "imageid" = "imageid")) #%>%
      #select(x, y, imageid, kmeans, cluster, metacluster)
    meta_mapped$metacluster[grepl("Artifact", meta_mapped$cluster)] <- "Artifact"
    colnames(meta_mapped)[colnames(meta_mapped) == "metacluster"] <- paste0("kmeans_metaclusters_", metacluster_range[i])
    df_list[[basename(j)]] <- meta_mapped
    
  }
  
  for (i in seq_along(df_list)) {
    file_name <- sub("scimap_clusters", "kmeans_metaclusters", names(df_list[i]))

    fwrite(df_list[[i]], file_name)
  }
}

```








