---
title: "Quality Control for MxIF"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged 
    geometry: margin=2cm
    highlight: textmate
    theme: journal
    fig_crop: false
    toc: true
    toc_float: true
  pdf_document: default
---
<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

requiredPackages <- c("ggplot2", "data.table", "dplyr")
for (package in requiredPackages) { #Installs packages if not yet installed
  if (!requireNamespace(package, quietly = TRUE))
  install.packages(package)
}

library(ggplot2)
library(data.table)
library(dplyr)

set.seed(123)

quantfile <- commandArgs(trailingOnly=T)[1]
roi <- tools::file_path_sans_ext(basename(quantfile))
configs_path <- commandArgs(trailingOnly=T)[2]
```

```{r}
## File paths
configs <- fread(configs_path)

cluster_outfile <- paste0("clusters_", roi, ".csv")

# QC parameters
sigsum_quantile_high <- as.numeric(configs$value[configs$object == "sigsum_quantile_high"])
sigsum_quantile_low <- as.numeric(configs$value[configs$object == "sigsum_quantile_low"])
bin_size <- as.numeric(configs$value[configs$object == "bin_size"])
density_cutoff <- as.numeric(configs$value[configs$object == "density_cutoff"])
allmarkers_outfile <- paste0("all_markers_clean_", roi, ".csv")
```

# ROI: `r roi`

# QC 

There are two main quality control steps:

1. SigSum: filtering outlier cells based on the sum of all marker signals per cell.
    - Cutoff: remove cells above the 99th percentile and below the 5th percentile
    
1. Low density cells
    - Remove "lonely" cells in low-density regions

```{r}
# Read in data
roi_df <- fread(quantfile)

# Reformat column names
if (length(grep("Cell:", colnames(roi_df))) == 0) {
  colnames(roi_df) <- sub(": ", ": Cell: ", colnames(roi_df)) # Insert "Cell:" into the column names for the marker genes for Seurat
}

colnames(roi_df) <- sub("\u00B5", "u", colnames(roi_df)) # replace mu with u
```

## Sigsum check
```{r}
# Extract the means for each marker
means_only_df <- roi_df %>% select(contains("Cell: Mean"))

# Get sum of all signals for each cell
means_only_df$sigsum <- rowSums(means_only_df)

# Calculate percentile cutoffs
sigsum_cutpoint_1 <- quantile(means_only_df$sigsum, sigsum_quantile_high)
sigsum_cutpoint_2 <- quantile(means_only_df$sigsum, sigsum_quantile_low)

# Add metric to df
roi_df$sigsum_metric = "Cell"
roi_df$sigsum_metric[means_only_df$sigsum > sigsum_cutpoint_1] <- "SIGSUM High"
roi_df$sigsum_metric[means_only_df$sigsum < sigsum_cutpoint_2] <- "SIGSUM Low"
roi_df$sigsum <- means_only_df$sigsum

n_sigsum_high = sum(roi_df$sigsum_metric == "SIGSUM High")
pct_sigsum_high = n_sigsum_high/nrow(roi_df)
n_sigsum_low = sum(roi_df$sigsum_metric == "SIGSUM Low")
pct_sigsum_low = n_sigsum_low/nrow(roi_df)
```

#### Cells flagged for high sigsum: `r n_sigsum_high` (`r round(pct_sigsum_high*100, 2)`%)

#### Cells flagged for low sigsum: `r n_sigsum_low` (`r round(pct_sigsum_low*100, 2)`%)

```{r}
# Plot distribution of sigsums with cutoffs
ggplot(data = means_only_df, aes(x = sigsum)) +
  geom_histogram() +
  theme_bw() +
  geom_vline(aes(xintercept = sigsum_cutpoint_1)) +
  geom_vline(aes(xintercept = sigsum_cutpoint_2)) +
  ggtitle(roi)
ggsave(paste0("sigsum_", roi, ".png"), device = "png", dpi = 300, width = 6, height = 5, units = "in")
```


## Bin density
```{r}
# Calculate bins for cell density
roi_df$binX <- floor(roi_df$`Centroid X`/bin_size)
roi_df$binY <- floor(roi_df$`Centroid Y`/bin_size)

# Count cells in each bin
bincounts <- roi_df %>% count(binX, binY) %>% rename(bin_density = n)

# Add bin density metric to df
roi_df <- roi_df %>% inner_join(bincounts) %>% mutate(low_bin_density = bin_density <= density_cutoff)

n_lowdensity = sum(roi_df$bin_density <= density_cutoff)
pct_lowdensity = n_lowdensity/nrow(roi_df)
```

#### Cells flagged for low density: `r n_lowdensity` (`r round(pct_lowdensity*100, 2)`%)

```{r}
# Plot cells with low density
ggplot(roi_df, aes(x = roi_df$`Centroid X`, y = roi_df$`Centroid Y`, fill = roi_df$low_bin_density)) +
  geom_point(pch=21,colour="white") +
  labs(x = "Centroid X", y = "Centroid Y", fill = "Low bin density") +
  scale_y_reverse() +
  theme_minimal() +
  ggtitle(roi)
ggsave(paste0("bin_density_", roi, ".png"), device = "png", dpi = 300, width = 6, height = 5, units = "in")
```


```{r}
## Instead of filtering, mark as artifacts here, and filter in clustering script
roi_df$qc <- "Cell"
roi_df$qc[(roi_df$sigsum > sigsum_cutpoint_1) | (roi_df$sigsum < sigsum_cutpoint_2)] <- "Artifact"
roi_df$qc[roi_df$bin_density <= density_cutoff] <- "Artifact"
roi_df$roi <- roi

cols_keep <- c("roi", 
               "sigsum_metric", "sigsum", "qc", "binX", "binY", "bin_density", 
               "low_bin_density")

roi_df <- roi_df %>% select(contains(c("Centroid X", "Centroid Y")), all_of(cols_keep), contains("Cell:"))
fwrite(roi_df, allmarkers_outfile)
```

